---
title: "Multiple Linear Regression"
author: "S. Wadgymar, N.M. Baran"
date: "10/27/2024"
output:
  pdf_document:
    latex_engine: xelatex
---
*adapted from OpenIntro biostats chapter 7 labs

In most practical settings, more than one explanatory variable is likely to be associated with a response. Multiple linear regression is an extension of simple linear regression that allows for more than one predictor variable in a linear model. As with simple linear regression, the response variable must be numerical, but the predictor variables can be either numerical or categorical. 

There are several applications of multiple regression. One of the most common applications in a clinical setting is estimating an association between a response variable and primary predictor of interest while adjusting for possible confounding variables.

First, we will explore multiple regression model by examining the possible association between cognitive function and the use of statins after adjusting for a potential confounding variable, age.

Reminder: Statins are a class of drug widely used to lower cholesterol. Research suggests that adults with elevated low density lipoprotein (LDL) cholesterol may be at risk for adverse cardiovascular events. A set of guidelines released in 2013 recommended statin therapy in individuals who are at high risk of adverse cardiovascular events, including individuals with Type II diabetes and moderately high LDL and non-diabetic individuals with atherosclerotic cardiovascular disease and high LDL. If these guidelines were to be followed, almost half of Americans ages 40 to 75 and nearly all men over 60 would be prescribed a statin.

However, some physicians have raised the question of whether treatment with a statin might be associated with an increased risk of cognitive decline.

This lab uses data from the Prevention of REnal and Vascular END-stage Disease (PREVEND) study. Clinical and demographic data for 4,095 individuals are stored in the PREVEND dataset in the oibiostat package. 

```{r}
library(oibiostat)
data(prevend.samp)
```

If the above code doesn't work, you need to install the oibiostat package using the following commands (minus the #)
```{r}
#install.packages("devtools")
#devtools::install_github("OI-Biostat/oi_biostat_data")
```


################################
1) Statin use is coded as an integer vector, where 0 represents a non-user and 1 represents a user. Run the code chunk below to convert the variable Statin into a factor variable, with levels NonUser and User. Describe what this code does below the code chunk (look at the data before and after running it).


```{r}
prevend.samp$Statin <- factor(prevend.samp$Statin, levels = c(0,1), labels = c("Nonuser", "User") )
```

A: This line of code recodes the integer dummy variable for statin use (0/1) and labels 0 to 'Nonuser' and 1 to 'User'.


################################
2) Adjust the code in the code chunk below to generate a scatterplot of RFFT score by Age.  Note that R will assign colors alphabetically to levels in alphabetical order.  Black comes before red alphabetically and Nonuser comes before User, therefore black represents Nonusers and red represents users.

```{r}
plot(prevend.samp$Age, prevend.samp$RFFT,
     pch = 19, col = prevend.samp$Statin, 
     xlab = "Age (yrs)",
     ylab = "RFFT Score",
     main = "")

legend('topright', legend=c("Nonuser", "User"), fill=c("black", "red"), bty='n')
```


Let's look at the correlation between Age and RFFT score.  Interpret the correlation coefficient below.

```{r}
cor(prevend.samp$Age, prevend.samp$RFFT)
```

A: This correlation describes a negative, moderate association between age and RFFT score. 

################################
3) Let's fit a simple regression model and interpret the slope coefficient. The lm function stands for linear model.  The arguments are Y~X.  Examine the output of the model and interpret all of its elements.  

```{r}
statin_lm <- lm(prevend.samp$RFFT~prevend.samp$Statin)
summary(statin_lm)
```

A: The first thing we should look at here is the F-statistic and p-value, which tells us that at least one of our explanatory variables (here we only have Statin use) is associated with the response variable (RFFT scores). 

The estimate for the intercept is 70.714 and it is statistically significant (p<0.05). The estimate for the effect of statin ue is -10.053 and this is significant, as well (P<0.05). If we build our mode in the form of y=b0 + b1(StatinUse), where b0 is the intercept and b1 is the slope of the line, the equation is:

y = 70.714-10.053(StatinUse). 

Thus the predicted RFFT score for a nonUser of statin is 70.714 -10.053 (0) = 70.714. The predicted RFFT score for a user of statin is 70.714-10.054(1) = 60.66. 

The adjusted R-squared is 0.02194, which means that 2.194% of the variation in RFFT scores can be explained by statin use. 


################################
4) Age is a potential confounder for the relationship between statin use and cognitive function. If older participants tend to use statins, and higher age is associated with lower cognitive ability, perhaps the observed negative association between cognitive ability and statin use is primarily driven by age.

Let's fit a multiple regression model for predicting RFFT score from statin use and age. Examine the output of the model and interpret all of its elements.  

```{r}
model.RFFTvsStatinAge = lm(prevend.samp$RFFT~prevend.samp$Statin + prevend.samp$Age)
summary(model.RFFTvsStatinAge)
```

Interpretation of intercept:The line of best fit intersects the Y-axis (i.e. when X=0) at 137.8822. This estimate is significant at P<0.05.

Interpretation of slope for Statin: The estimate for  

Interpretation of slope for Age:

Interpretation of R2:





################################
5) Let's check all assumptions of the linear model. Run the code below and address whether assumptions are met.

```{r}
#Extract residuals
hist(resid(model.RFFTvsStatinAge))

#Examine residuals (here, fitted values = predicted values)
#Just look at the first two plots for now
plot(model.RFFTvsStatinAge)

#Variance test of residuals against Statin use
var.test(resid(model.RFFTvsStatinAge)~prevend.samp$Statin)
```

A:





################################
6) How does the simple linear model compare to the multiple regression?  Answer the questions below:

a) The adjusted R2 value accounts for the increased number of predictors and should be compared across models to determine the best model. How does the adjusted R2 of the simple linear model compare with the multiple linear model? Which do you think is a better model for the data?

A:




b) As in simple linear regression, inferences can be made about the slope parameters estimated by the model slope coefficients. Based on the multiple regression model, is there a statistically significant association between RFFT score and statin use? 

A: 





c) In a clinical setting, the interpretive focus lies on reporting the nature of the association between the primary predictor and the response, while specifying which potential confounders have been adjusted for. Briefly respond to a clinician who is concerned about a possible association between statin use and decreased cognitive function, based on the analyses conducted in this lab.

A: 

        
        
```{r}
model.RFFTxStatinAge = lm(prevend.samp$RFFT~prevend.samp$Statin + prevend.samp$Age+prevend.samp$Statin*prevend.samp$Age)
summary(model.RFFTxStatinAge)
```        
```{r}
model.RFFTxStatinAge = lm(prevend.samp$RFFT~prevend.samp$Statin*prevend.samp$Age)
summary(model.RFFTxStatinAge)
```  

################################
7) Let's add the lines of best fit predicted by the models to the graph.

First, let's identify the slopes and intercepts.  The coef function will isolate the coefficients (parameters) from the model.  There are three parameters from this model.  You can list individual parameters by specifying [1], [2], or [3] at the end of the code (I do that in the next code chunk).

```{r}
coef(model.RFFTvsStatinAge)
```


Next, let's add some trend lines.  The abline(a,b) function will let you specify the intercept (a) and the slope (b) for a line.  I wrote the function to graph the line for Statin non-users.  You need to adjust the code to graph the line for Statin users.  Note that you have to run the entire code chunk to see if the lines are working (you can't just run the abline code by itself).

```{r}
##Paste the code for the graph you previously made

plot(prevend.samp$Age, prevend.samp$RFFT,
     pch = 19, col = prevend.samp$Statin, 
     xlab = "Age (yrs)",
     ylab = "RFFT Score",
     main = "")


##line for Statin non-users
abline(a=coef(model.RFFTvsStatinAge)[1], 
       b=coef(model.RFFTvsStatinAge)[3],
       lwd=2)

##line for Statin users
abline(a=coef(model.RFFTvsStatinAge)[1], 
       b=coef(model.RFFTvsStatinAge)[3],
       lwd=2, lty=2, col="red")

```


















2. Explore the relationship between RFFT score and statin use with the data in prevend.samp

a) Statin use is coded as an integer vector, where 0 represents a non-user and 1 represents a user. Run the code chunk below to convert the variable Statin into a factor variable, with levels NonUser and User. Talk through what this code does (look at the data before and after running it).

```{r}
prevend.samp$Statin <- factor(prevend.samp$Statin, levels = c(0,1), labels = c("Nonuser", "User") )
```


b) Create a plot showing the association between RFFT score and statin use. Describe what you see.

Since statin use is a categorical variable, we know that the relationship is linear (the relationship between any two points is linear). Thus, it is best to look at the data with a boxplot:

```{r}
boxplot(prevend.samp$RFFT~prevend.samp$Statin)
```

It looks like there is possibly a relationship between statin use and lower (negative) RFFT score.

c) Fit a simple regression model and interpret the slope coefficient.

```{r}
statin_lm <- lm(prevend.samp$RFFT~prevend.samp$Statin)
summary(statin_lm)
```

The linear model is:
RFFT = 70.714 + (-10.053)*Statin_use

This suggests that if a person uses statin then their RFFT score decreases by about -10 points. 

d) Discuss whether the model from part c) is sufficient for understanding whether statin use is associated with decreased cognitive ability.

Although the p-value for the slope is very small, we still likely want to be cautious about this relationship as the decline in age could simply be due to those people that take statin tend ot be older. 

Age is a potential confounder for the relationship between statin use and cognitive function. If older participants tend to use statins, and higher age is associated with lower cognitive ability, perhaps the observed negative association between cognitive ability and statin use is primarily driven by age.

4. Fit a multiple regression model for predicting RFFT score from statin use and age. How does this linear model code differ from the simple linear regression above? 

```{r}
model.RFFTvsStatinAge = lm(prevend.samp$RFFT~prevend.samp$Statin + prevend.samp$Age)
summary(model.RFFTvsStatinAge)
```
After the '~' there are two predictors.

a) Write the equation of the linear model.
RFFT = .8509*statin_use + -1.271*age + 137.882

b) As with simple linear regression we need to check assumptions before interpreting our model. Check all assumptions of the linear model. (hint: to check for linearity assess the residuals against each of the predictors. To check for constant variance check the residuals against the response variable.)

Pull out the residuals and save them:
```{r}
statin_age_lm_residuals = model.RFFTvsStatinAge$residuals
```

Does there seem to be a linear relationship between each of the predictor variables and the response variable? Since we cannot plot three variables at a time on a scatterplot we can look at the residuals compared to each predictor variable to see if there is a linear relationship:

residuals vs age:
```{r}
plot(prevend.samp$Age, statin_age_lm_residuals)
```
There are no apparent trends; the data scatter evenly above and below the horizontal
line. There does not seem to be remaining nonlinearity with respect to age after the
model is fit.

It is not necessary to assess linearity with respect to statin use since statin use is measured as a categorical variable. A line drawn through two points (that is, the mean of the two groups defined by a binary variable) is necessarily linear.

We also want to look for constant variance in the residuals as compared to the predicted values of the reponse variable (and I have included extra formatting to make the plot pretty):
```{r}
plot(statin_age_lm_residuals ~ fitted(model.RFFTvsStatinAge),
main = "Resid. vs Predicted RFFT in PREVEND (n = 500)",
xlab = "Predicted RFFT Score", ylab = "Residual",
pch = 21, col = "cornflowerblue", bg = "slategray2",
cex = 0.60)
abline(h = 0, col = "red", lty = 2)
```
Are the observations independent?
It is reasonable to assume that the observations in this dataset are independent. The participants were recruited from a large city in the Netherlands for a study focusing on factors associated with renal and cardiovascular disease.

Assess whether the residuals are approximately normally distributed.
```{r}
qqnorm(statin_age_lm_residuals,
pch = 21, col = "cornflowerblue", bg = "slategray2", cex = 0.75,
main = "Q-Q Plot of Residuals")
qqline(statin_age_lm_residuals, col = "red", lwd = 2)
```
The residuals are reasonably normally distributed, with only slight departures from
normality in the tails.

c) Interpret the slope coefficient for statin use. Compare the coefficient to the one from the simple regression model between RFFT score and statin use.
If age is kept constant than the RFFT score increases by .8509 for someone who uses statin compared to someone that doesn't.

d) Interpret the slope coefficient for age.
If statin use is kept constant, then the RFFT score is expected to decrease by 1.271 as age increases by a year.

e) The adjusted R2 value accounts for the increased number of predictors and should be compared across models to determine the best model. How does the adjusted R2 of the simple linear model compare with the multiple linear model? Which do you think is a better model for the data?
The adjusted R2 for the simple linear regression model was .02194 and the adjusted R2 for the multiple linear regression is .2823. So the multiple linear regression explains about ten times as much of the change in RFFT score with these two predictors. Thus, the multiple regression model is likely the better model. 

f) As in simple linear regression, inferences can be made about the slope parameters estimated by the model slope coefficients. Based on the multiple regression model, is there a statistically significant association between RFFT score and statin use? 
No. Once age is accounted for there is not enough evidence to reject the null hypothesis that there is no association (beta = 0) between statin use and RFFT.

g) In a clinical setting, the interpretive focus lies on reporting the nature of the association between the primary predictor and the response, while specifying which potential confounders have been adjusted for. Briefly respond to a clinician who is concerned about a possible association between statin use and decreased cognitive function, based on the analyses conducted in this lab.
Once you adjust for the confounding variable of age, there is no evidence to suggest that statin use is associated with a lower RFFT score. Specifically after adjusting for age, the estimated difference in mean RFFT score between statin users and nonusers is .85 points; there is a 74% chance of observing such a difference if there is no difference between mean RFFT score in the population of statin users and nonusers.
    

        
        
        
        
        
